Efficient Calculation of Regularized Discriminant Analysis for High-Dimensional Classification

### 2013-12-12 reports/paper-hdrda.Rnw
Updated LaTeX page dimensions.
Reorganized so that all the notation is at the beginning of Section 2.

@@

\subsection{Preliminaries}
In discriminant analysis we wish to assign correctly an unlabeled
$p$-dimensional observation vector $\bm x$ to one of $K$ unique, known classes
(or populations) by constructing a classifier from $N$ training
observations. Let $\bm x_i = (x_{i1}, \ldots, x_{ip}) \in \mathbb{R}_{p \times
  1}$ be the $i$th observation $(i = 1, \ldots, N)$ with true, unique membership
$y_i \in \{\omega_1, \ldots, \omega_K\}$, where $\mathbb{R}_{a \times b}$
denotes the matrix space of all $a \times b$ matrices over the real field
$\mathbb{R}$. We assume that $(\bm x_i, y_i)$ is a realization from a mixture
distribution $p(\bm x) = \sum_{k=1}^K p(\bm x | \omega_k) p(\omega_k)$, where
$p(\bm x | \omega_k)$ is the probability density function of the $k$th class and
$p(\omega_k)$ is the prior probability of class membership of the $k$th
class. We further assume that each of $p(\omega_k)$ is equal, $k = 1, \ldots, K$.

Quadratic discriminant analysis is the optimal Bayesian decision rule with
respect to a $0-1$ loss function when $p(\bm x | \omega_k)$ is the probability
density function of the multivariate normal distribution with known mean vectors
$\bm \mu_k \in \mathbb{R}_{p \times 1}$ and known covariance matrices $\bm
\Sigma_k \in \mathbb{R}_{p \times p}^{>}$, $k = 1, 2, \ldots, K$, where
$\mathbb{R}_{p \times p}^{>}$ denotes the cone of real $p \times p$
positive-definite matrices. Because $\bm \mu_k$ and $\bm \Sigma_k$ are typically
unknown, we assign an unlabeled observation $\bm x$ to class $\omega_k$ with the
sample quadratic discriminant analysis classifier
\begin{align}
	D_{QDA}(\bm x) = \argmin_{k} (\bm x - \xbar_k)^{T}\widehat{\bm
    \Sigma}_k^{-1}(\bm x - \xbar_k) + \log |\widehat{\bm \Sigma}_k|, \label{eq:qda}
\end{align}
where $\xbar_k$ and $\widehat{\bm \Sigma}_k$ are the maximum likelihood
estimators of $\bm \mu_k$ and $\bm \Sigma_k$, respectively, and $\bm A^{T}$
denotes the matrix transpose of $\bm A \in \mathbb{R}_{m \times p}$. If we
assume further that $\bm \Sigma_k = \bm \Sigma$, $k = 1, \ldots, K$, then the
pooled sample covariance matrix $\widehat{\bm \Sigma}$ is substituted for
$\widehat{ \bm \Sigma}_k$ in \eqref{eq:qda}, where
\begin{align}
	\widehat{\bm \Sigma} = N^{-1} \sum_{k=1}^K n_k \widehat{\bm \Sigma}_k \label{eq:pooled-cov}
\end{align}
is the maximum likelihood estimator for $\bm \Sigma$. In this case \eqref{eq:qda}
reduces to the sample linear discriminant analysis classifier, and the
log-determinant is omitted because it is constant across all classes.

The smallest eigenvalues and the directions associated with their eigenvectors
highly influence \eqref{eq:qda}. In fact, the eigenvalues of $\widehat{ \bm
  \Sigma}_k$ are well-known to be biased if $p \ge n_k$ such that the smallest
eigenvalues are underestimated \citep{Seber:2004uh}. Moreover, if $p > n_k$,
then rank$(\widehat{\bm \Sigma}_k) \le n_k$, which implies that at least $p -
n_k$ eigenvalues of $\widehat{\bm \Sigma}_k$ are zero. Furthermore, although
more feature information is available to discriminate among the $K$ classes, if
$p > n_k$, \eqref{eq:qda} is incalculable because $\widehat{\bm \Sigma}_k^{-1}$
does not exist.

@@

Throughout the remainder of this document we also require the following
additional notation. Let $\bm I_m$ be the $m \times m$ identity matrix, and let
$\bm 0_{m \times p}$ be the $m \times p$ matrix of zeros, such that $\bm 0_m$ is
understood to denote $\bm 0_{m \times m}$. For $c \in \mathbb{R}$, let $c^+ =
1/c$ if $c \ne 0$ and $0$ otherwise. Let $\bm A \oplus \bm B$ denote the direct
sum of $\bm A \in \mathbb{R}_{r \times r}$ and $\bm B \in \mathbb{R}_{s \times
  s}$ \citep[Chapter 1]{Lutkepohl:1996uz}. Let $\bm A^+$ and $\mathcal{N}(\bm A)
= \{\bm z \in \mathbb{R}_{p \times 1} : \bm A \bm z = \bm 0_{m \times 1}\}$
denote the Moore-Penrose pseudoinverse and null space of $\bm A \in
\mathbb{R}_{m \times p}$, respectively. Let $V^{\perp} = \{\bm y \in
\mathbb{R}_{p \times 1} : \bm y^{T} \bm v = 0$ for all $\bm v \in V \}$ denote
the orthogonal complement of a vector space $V \subset \mathbb{R}_{p \times 1}$.

>>>>

\pagestyle{plain}
%----------------Page dimensions ----------------
\oddsidemargin 0.0in
\evensidemargin 0.0in
\topmargin -0.75in
\leftmargin 0in
\headheight 0.0in
\headsep 0.5in
\footskip 0.5in
\footnotesep 0.0in
\textwidth 6.5in
\textheight 9.5in

@@

\subsection{Preliminaries}
To facilitate our discussion of covariance-matrix regularization and regularized
discriminant analysis, we require the following notation. Let $\mathbb{R}_{a
  \times b}$ denotes the matrix space of all $a \times b$ matrices over the real
field $\mathbb{R}$. Denote by $\bm I_m$ the $m \times m$ identity matrix, and
let $\bm 0_{m \times p}$ be the $m \times p$ matrix of zeros, such that $\bm
0_m$ is understood to denote $\bm 0_{m \times m}$. Let $\bm A \oplus \bm B$
denote the direct sum of $\bm A \in \mathbb{R}_{r \times r}$ and $\bm B \in
\mathbb{R}_{s \times s}$ \citep[Chapter 1]{Lutkepohl:1996uz}. Let $\bm A^{T}$,
$\bm A^+$, and $\mathcal{N}(\bm A) = \{\bm z \in \mathbb{R}_{p \times 1} : \bm A
\bm z = \bm 0_{m \times 1}\}$ denote the transpose, the Moore-Penrose
pseudoinverse, and the null space of $\bm A \in \mathbb{R}_{m \times p}$,
respectively. Let $\mathbb{R}_{p \times p}^{>}$ denote the cone of real $p
\times p$ positive-definite matrices. Similarly, let $\mathbb{R}_{p \times
  p}^{\ge}$ denote the cone of real $p \times p$ positive-semidefinite
matrices. Let $V^{\perp} = \{\bm y \in \mathbb{R}_{p \times 1} : \bm y^{T} \bm v
= 0$ for all $\bm v \in V \}$ denote the orthogonal complement of a vector space
$V \subset \mathbb{R}_{p \times 1}$. For $c \in \mathbb{R}$, let $c^+ = 1/c$ if
$c \ne 0$ and $0$ otherwise.

In discriminant analysis we wish to assign correctly an unlabeled
$p$-dimensional observation vector $\bm x$ to one of $K$ unique, known classes
(or populations) by constructing a classifier from $N$ training
observations. Let $\bm x_i = (x_{i1}, \ldots, x_{ip}) \in \mathbb{R}_{p \times
  1}$ be the $i$th observation $(i = 1, \ldots, N)$ with true, unique membership
$y_i \in \{\omega_1, \ldots, \omega_K\}$. We assume that $(\bm x_i, y_i)$ is a
realization from a mixture distribution $p(\bm x) = \sum_{k=1}^K p(\bm x |
\omega_k) p(\omega_k)$, where $p(\bm x | \omega_k)$ is the probability density
function of the $k$th class and $p(\omega_k)$ is the prior probability of class
membership of the $k$th class. We further assume that each of $p(\omega_k)$ is
equal, $k = 1, \ldots, K$.

Quadratic discriminant analysis is the optimal Bayesian decision rule with
respect to a $0-1$ loss function when $p(\bm x | \omega_k)$ is the probability
density function of the multivariate normal distribution with known mean vectors
$\bm \mu_k \in \mathbb{R}_{p \times 1}$ and known covariance matrices $\bm
\Sigma_k \in \mathbb{R}_{p \times p}^{>}$, $k = 1, 2, \ldots, K$. Because $\bm
\mu_k$ and $\bm \Sigma_k$ are typically unknown, we assign an unlabeled
observation $\bm x$ to class $\omega_k$ with the sample quadratic discriminant
analysis classifier
\begin{align}
	D_{QDA}(\bm x) = \argmin_{k} (\bm x - \xbar_k)^{T}\widehat{\bm
    \Sigma}_k^{-1}(\bm x - \xbar_k) + \log |\widehat{\bm \Sigma}_k|, \label{eq:qda}
\end{align}
where $\xbar_k$ and $\widehat{\bm \Sigma}_k$ are the maximum likelihood
estimators of $\bm \mu_k$ and $\bm \Sigma_k$, respectively. If we assume further
that $\bm \Sigma_k = \bm \Sigma$, $k = 1, \ldots, K$, then the pooled sample
covariance matrix $\widehat{\bm \Sigma}$ is substituted for $\widehat{ \bm
  \Sigma}_k$ in \eqref{eq:qda}, where
\begin{align}
	\widehat{\bm \Sigma} = N^{-1} \sum_{k=1}^K n_k \widehat{\bm \Sigma}_k \label{eq:pooled-cov}
\end{align}
is the maximum likelihood estimator for $\bm \Sigma$. In this case \eqref{eq:qda}
reduces to the sample linear discriminant analysis classifier, and the
log-determinant is omitted because it is constant across the $K$ classes.

The smallest eigenvalues and the directions associated with their eigenvectors
can highly influence \eqref{eq:qda}. In fact, the eigenvalues of $\widehat{ \bm
  \Sigma}_k$ are well-known to be biased if $p \ge n_k$ such that the smallest
eigenvalues are underestimated \citep{Seber:2004uh}. Moreover, if $p > n_k$,
then rank$(\widehat{\bm \Sigma}_k) \le n_k$, which implies that at least $p -
n_k$ eigenvalues of $\widehat{\bm \Sigma}_k$ are zero. Furthermore, although
more feature information is available to discriminate among the $K$ classes, if
$p > n_k$, \eqref{eq:qda} is incalculable because $\widehat{\bm \Sigma}_k^{-1}$
does not exist.

@@
